A Step-Indexed Logic in Agda

### 2023-07-06 papers/cpp2024/paper.tex
check

\section{Introduction}

@@

>>>>

\section{Introduction}
\label{sec:intro}

@@

Our goal is to define an operator for recursive predicates and relations
with syntax that is something like $μᵒ x. R$, where $x$ is the name of the
recursive relation and $R$ is the definition of the relation, which
can refer to $x$. We shall prove a fixpoint theorem which states that
the recursive predicate is equal to its unfolding, something like the
following.
\[
  (μᵒ x. R) \app δ \app a ≡ᵒ R \app δ(x:= μᵒ x. R) \app a
\]
where $δ$ is an environment mapping variables to predicates.

Our proof of this fixpoint theorem is inspired by the fixpoint theorem
of \citet{Appel:2001aa}. In that work, \citet{Appel:2001aa} use
step-indexing to give a semantic definition of recursive types. Their
fixpoint theorem proves that a recursive type is equal to its unfolding.
They define a (semantic) type $\tau$ to be a relation between step indexes and
syntactic values. They do not define a syntax for types, but instead
define operators for constructing semantic types as follows.
\begin{align*}
  ⊥ &= \{ \} \\
  ⊤ &= \{ ⟨k,v⟩ \mid k ∈ ℕ\} \\
  \mathbf{int} &= \{⟨k,n⟩ \mid k ∈ ℕ, n ∈ ℤ \}\\
  τ₁ × τ₂ &= \{ ⟨k,(v₁,v₂)⟩ \mid ∀j<k. ⟨j,v₁⟩∈τ₁, ⟨j,v₂⟩∈τ₂ \} \\
  τ₁ → τ₂ &= \{ ⟨k,λx.e⟩ \mid ∀j<k.∀v. ⟨j,v⟩∈τ₁ ⇒ e[v/x] :ⱼ τ₂ \} \\
  μ F &= \{ ⟨k,v⟩ \mid ⟨k,v⟩ ∈ F^{k\plus 1}(⊥) \}
    & \text{if } F : \tau \to \tau'
\end{align*}
Their fixpoint theorem says that for any well founded $F$,
\[
  μ F = F(μF)
\]
A well founded function $F$ on types is
one in which each pair in the output $⟨k,v⟩$ only depends
on later pairs in the input, that is, pairs of the form $⟨j,v′⟩$
where $j < k$. \citet{Appel:2001aa} characterize this dependency
with the help of the $k$-approximation function:
\[
  \kapprox(k,τ) = \{ ⟨j,v⟩ \mid j < k, ⟨j,v⟩ ∈ τ\} 
\]
They define a \emph{well founded} function $F$ to be one that
satisfies the following equation.
\[
  \kapprox(k \plus 1, F(τ)) = \kapprox(k \plus 1, F(\kapprox(k,τ)))
\]

Functions over semantic types are not always well founded.  For
example, the identity function $λα.α$ is not well founded, so one
cannot apply the fixpoint theorem to the recursive type $μ(λα.α)$
(which corresponds to the syntactic type $μα.α$).
On the other hand, the function
$λα.α×α$ is well founded because of the strict less-than in the
definition of the $×$ operator. So the fixpont theorem applies to
$μ(λα.α×α)$.  In general, a function built from the type operators is
well founded so long as the recursive $α$ only appears underneath a
type constructor such as $×$ or $→$. \citet{Appel:2001aa} prove this
fact, which relies on the auxilliary notion of a nonexpansive
function. In such a function, the output can depend on pairs at the
current step index as well as later ones. So a \emph{nonexpansive}
function satisfies the following equation.
\[
  \kapprox(k, F(τ)) = \kapprox(k, F(\kapprox(k,τ)))
\]
For example, $λα.α$ is nonexpansive and so is $λα.\mathbf{int}$.
\citet{Appel:2001aa} then prove lemmas about the type constructors.
For example, regarding products, they prove that if $F$ and $G$
are nonexpansive functions, then so is $λ α. (F α) × (G α)$.

It is worth noting that \citet{Appel:2001aa} neglect to prove such 
lemmas for the $μ$ operator itself. For example, given $F : τ₁ → τ₂ → τ₃$
that is nonexpansive in its first parameter and well founded in
its second, then $λ α. μ (F α)$ is nonexpansive.
On the other hand, if $F$ is well founded in both parameters,
then $λ α. μ (F α)$ is well founded. We shall return to this point later.

Comparing the type operators of \citet{Appel:2001aa} to the logic
operators of SIL, there are striking similarities. The function type
operator is quite similar to implication, although one difference is
that the function type operator uses strict less-than whereas
implication uses non-strict less-than. The logic introduces the
``later`` operator, whereas the type operators essentially bake the
later operator into the type operators through their use of strict
less-than.

Our definition of recursive predicates will be similar to the
recursive type of \citet{Appel:2001aa} in that we shall define the
meaning of a recursive predicate by iteration.  On the other hand, we
do not want a fixpoint theorem that requires the user of the logic to
provide a proof that a particular proposition is well
founded. Instead, we shall introduce a type system for propositions
that ensure that $μ$ is only applied to well founded propositions, and
that the proof of well foundedness is provided by our logic operators,
not by the user of the logic.
